---
layout: page
title: Save the data in plain text files
description: "Data organization: Use plain text files"
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, results="asis")
library(broman)
```

Keep a copy of your data files in a plain text format, with comma or
tab delimiters.  I generally use comma-delimited (CSV) files.

So a spreadsheet like this:

```{r example_rectangle}
set.seed(20150803)
mat <- data.frame(id=101:105,
                  sex=sample(c("Male", "Female"), 5, replace=TRUE),
                  glucose=myround(runif(5, 70, 150), 1),
                  insulin=myround(runif(5, 0.5, 1.5), 2),
                  triglyc=myround(runif(5, 70, 300), 1),
                  stringsAsFactors=FALSE)

excel_fig(mat, direct2svg=TRUE)
```

Would become a plain text file like this:

    id,sex,glucose,insulin,triglyc
    101,Male,134.1,0.60,273.4
    102,Female,120.0,1.18,243.6
    103,Male,124.8,1.23,297.6
    104,Male,83.1,1.16,142.4
    105,Male,105.2,0.73,215.7

It's not pretty to look at, but you can open the file in Excel and it
will look pretty again. More importantly, this sort of non-proprietary file
format does not and never will require any sort of special software.

(And most of the time, the first thing I do with an Excel file is
convert it into a set of CSV files, one per worksheet, because they're
easier to handle in code.)

If any of the cells in your data include commas, Excel will put
double-quotes around the contents of each cell when it's saved in CSV
format. That requires slightly more finesse to deal with, but it's
generally not a concern.

To save an Excel file as a comma-delimited file:

- From the menu bar, File → Save As
- Next to "Format:", click the drop-down menu and select "Comma Separated Values (CSV)"
- Click "Save"
- Excel will say something like, "This workbook contains features that will
  not work...". Ignore that and click "Continue".
- Quit Excel. It will ask you, "Do you want to save the changes you
  made?" Click "Don't Save", because you just saved them. (Excel
  really doesn't want you to use a format other than it's own.)

Note that there's also an option to save as "Tab Delimited Text". Many
people prefer that. (I don't like tab delimiters so much, myself, as
multiple adjacent tab characters are hard to discern by eye.)

Personally, I prefer to use a vertical bar (`|`) as a delimiter
between fields, because the data itself will seldom contain a vertical
bar. But you can't get that from Excel.

---

Next up: [Other things to avoid](avoid.html).

(Previous: [Use data validation to avoid data entry mistakes](validation.html).)

<style>
svg { font-family: sans-serif; }
svg text { font-family: sans-serif; }
</style>
