---
layout: page
title: Make it a rectangle
description: "Data organization: Use a rectangular layout"
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=2.5,
                      fig.path="Figs/no_empty_cells-",
                      dev="svg", echo=FALSE)
library(broman)
```


The best layout for your data within in a spreadsheet is as a big
rectangle with rows corresponding to the subjects and columns
corresponding to variables. The first row should contain variable
names. (*Please don't use more than one row for the variable names.*)
Here's an example:

```{r example_rectangle, fig.height=2}
set.seed(20150803)
mat <- data.frame(id=101:105,
                  sex=sample(c("Male", "Female"), 5, replace=TRUE),
                  glucose=myround(runif(5, 70, 150), 1),
                  insulin=myround(runif(5, 0.5, 1.5), 2),
                  triglyc=myround(runif(5, 70, 300), 1),
                  stringsAsFactors=FALSE)

excel_fig(mat)
```

Some data sets won't fit nicely into a rectangle, but they'll usually
fit into a set of rectangles, in which you can make a set of Excel files,
each with a rectangle of data. Some data don't even fit into a set of
rectangles, but then maybe spreadsheets are not the best format for
them, as spreadsheets seem inherently rectangular.

The data files that I receive are not very often in rectangular
form. More often, there seem to be bits of data sprinkled about.
Like this:

```{r bad_example_1, fig.height=2.8}
mat2 <- t(mat)

mat2 <- cbind(rownames(mat2), mat2)
mat2 <- rbind(rep("", 6),
              mat2[1:2,],
              rep("", 6),
              mat2[c(1,3),],
              rep("", 6),
              mat2[c(1,4),])
mat2[c(2,5,8),1] <- ""

excel_fig(mat2, col_names=FALSE)
```

Or maybe like this:

```{r bad_example_2, fig.height=2.8, fig.width=6}
mat3 <- rbind(c("1MIN", rep("", 6)),
              c("", "", "Normal", "", "", "Mutant", ""),
              c("B6", 146.6, 138.6, 155.6, 166.0, 179.3, 186.9),
              c("BTBR", 245.7, 240.0, 243.1, 177.8, 171.6, 188.1),
              rep("", 7),
              c("5MIN", rep("", 6)),
              c("", "", "Normal", "", "", "Mutant", ""),
              c("B6", 333.6, 353.6, 408.8, 450.6, 474.4, 423.8),
              c("BTBR", 514.4, 610.6, 597.9, 412.1, 447.4, 446.5))

excel_fig(mat3, col_names=FALSE)
```

In both of these cases, the data analyst has to study the layout, figure out
what everything means, and then spend some time rearranging things.

Another recent example: a separate worksheet for each subject, each in
a complicated format like this:

```{r bad_example_3, fig.height=3.5, fig.width=8}
mat4 <- rbind(rep("", 7),
              c("Date", "11/3/14", rep("", 5)),
              c("Days on diet", "126", rep("", 5)),
              c("Mouse #", "43", rep("", 5)),
              c("sex", "f", rep("", 5)),
              c("experiment", "", "values", "", "", "mean", "SD"),
              c("control", "", 0.186, 0.191, 1.081, "", ""),
              c("treatment A", "", 7.414, 1.468, 2.254, "", ""),
              c("treatment B", "", 9.811, 9.259, 11.296, "", ""),
              rep("", 7),
              c("fold change", "", "values", "", "", "mean", "SD"),
              c("treatment A", "", 7.414, 1.468, 2.254, "", ""),
              c("treatment B", "", 9.811, 9.259, 11.296, "", ""))

for(i in 7:9) {
    # calc mean and SD
    x <- as.numeric(mat4[i, 3:5])
    mat4[i,6] <- round(mean(x), 2)
    mat4[i,7] <- round(sd(x), 2)

    if(i > 7) {
      # fold change
      y <- as.numeric(mat4[7, 3:5])
      z <- x/mean(y)
      mat4[i+4,3:5] <- round(z, 2)
      mat4[i+4,6] <- round(mean(z), 2)
      mat4[i+4,7] <- round(sd(z), 2)
    }
}
excel_fig(mat4)
```

If all of the worksheets have exactly the same layout, then it's not
too hard to pull out the relevant information and combine it into a
rectangle (I'd write a [python](https://www.python.org) or
[ruby](https://www.ruby-lang.org) script). But I prefer to not have
means and SDs and fold change calculations cluttering things up, and
it seems that even for data entry, it'd be easier to have all of the
measurements on one worksheet.

Sometimes it's hard to see how to reorganize things as a
rectangle. Consider this example:

```{r bad_example_4, fig.width=8, fig.height=5}
mat5 <- rbind(c("", "GTT date", "GTT weight", "time", "glucose mg/dl", "insulin ng/ml"),
              c("321", "2/9/15", "24.5",  0,  99.2, "lo off curve"),
              c("", "", "",               5, 349.3, 0.205),
              c("", "", "",              15, 286.1, 0.129),
              c("", "", "",              30, 312.0, 0.175),
              c("", "", "",              60,  99.9, 0.122),
              c("", "", "",             120, 217.9, "lo off curve"),
              c("322", "2/9/15", "18.9",  0, 185.8, 0.251),
              c("", "", "",               5, 297.4, 2.228),
              c("", "", "",              15, 439.0, 2.078),
              c("", "", "",              30, 362.3, 0.775),
              c("", "", "",              60, 232.7, 0.500),
              c("", "", "",             120, 260.7, 0.523),
              c("323", "2/9/15", "24.7",  0, 198.5, 0.151),
              c("", "", "",               5, 530.6, "off curve lo"))

excel_fig(mat5, col_names=FALSE)
```

Well, it's sort of a rectangle; we could just fill in the empty cells,
like this:

```{r bad_example_4_filled, fig.width=8, fig.height=5}
mat5[1,1] <- "id"
for(i in 1:3) {
    mat5[3:7, i] <- mat5[2,i]
    mat5[(3:7)+6, i] <- mat5[8,i]
    mat5[15, i] <- mat5[14,i]
}

excel_fig(mat5, col_names=FALSE)
```

But it seems wrong to repeat the weights like that. It's better to
make two separate tables, one with the weights, and one with these
other measurements (which are for in vivo assay, the glucose tolerance
test: give a mouse some glucose and measure serum glucose and insulin
levels at different times afterwards).

So you'd have one table with the weights, like this:

```{r bad_example_4_weights, fig.width=6/5*4, fig.height=5/16*5}
mat5wt <- mat5[c(1,2,8,14), 1:3]
excel_fig(mat5wt, col_names=FALSE)
```

And then another rectangle with the GTT results, like this:

```{r bad_example_4_gtt, fig.width=6, fig.height=5}
mat5gtt <- mat5[,c(1, 4:6)]
mat5gtt[1,2] <- "GTT time"
excel_fig(mat5gtt, col_names=FALSE)
```

---

Next up: [Create a data dictionary](dictionary.html).
