---
layout: page
title: No empty cells
description: "Data organization: No empty cells"
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=2.5,
                      fig.path="Figs/no_empty_cells-",
                      dev="svg", echo=FALSE)
library(broman)
```

Fill in all cells. Use some common code for missing data.
Not everyone agrees with me (for example,
[White et al. (2013)](http://library.queensu.ca/ojs/index.php/IEE/article/view/4608/4898)
state a preference for leaving cells blank), but I'd prefer to have
"`NA`" or even a hyphen in the cells with missing data, to make sure
it's clear that the data are known to be missing rather than
unintentially left blank.

I also often see cells left blank when a single value is meant to be
repeated multiple times. For example, one might put the date in only a
few cells, like this:

```{r skipping_cells}
mat <- data.frame(id=101:107,
                  date=c("2015-06-14", "", "2015-06-18", "", "", "2015-06-20", ""),
                  glucose=myround(c(149.3, 95.3, 97.5, 117.0, 108.0, 149.0, 169.4), 1),
                  stringsAsFactors=FALSE)
excel_fig(mat)
```

**Don't do that!** If the rows were to be sorted at some point, that
date column would be completely mangled.

It's much better to fill them all in, like this:

```{r skipping_cells_fixed}
mat$date[2] <- mat$date[1]
mat$date[4:5] <- mat$date[3]
mat$date[7] <- mat$date[6]
excel_fig(mat)
```

I also see this sort of thing for information about different
treatments. For example, I recently saw a file like the following:

```{r more_skipping_cells, fig.width=8, fig.height=1.5}
mat2 <- data.frame(c("", "strain", "A", "B"),
                   c("1 min", "normal", "147", "246"),
                   c("", "", "139", "240"),
                   c("", "mutant", "166", "178"),
                   c("", "", "179", "172"),
                   c("5 min", "normal", "334", "514"),
                   c("", "", "354", "611"),
                   c("", "mutant", "451", "412"),
                   c("", "", "474", "447"),
                    stringsAsFactors=FALSE)

excel_fig(mat2, col_names=FALSE)
```

We'll [talk more about layout shortly](rectangle.html), but while it's sort
of clear, to a human, what's intended above, the computer will have a
hard time with it.

You could fill in some of those cells, to make it more clear, but even
better would be to make a "tidy" version of the data
([more on what is meant by "tidy" later](rectangle.html), as part of
the discussion of layout), with each row being one replicate, as follows:

```{r more_skipping_cells_fixed, fig.height=4, fig.width=5/4*6}
mat2b <- data.frame(strain=rep(c("A", "A", "B", "B"), 4),
                    genotype=rep(rep(c("normal", "mutant"), each=4),2),
                    min=rep(c(1, 5), each=8),
                    replicate=rep(c(1,2),8),
                    response=c("147", "139", "246", "240",
                               "166", "179", "178", "172",
                               "334", "354", "514", "611",
                               "451", "474", "412", "447"),
                    stringsAsFactors=FALSE)

excel_fig(mat2b)
```

**No empty cells!**

---

Next up: [Put just one thing in a cell](one_thing_per_cell.html).
