---
layout: page
title: Don't use font color or highlighting as data
description: "Data organization: Don't use font color or highlighting as data."
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=2.5,
                      fig.path="Figs/no_highlighting-",
                      dev="svg", echo=FALSE)
library(broman)
```

You might be tempted to highlight particular cells with suspicious
data, or rows that should be ignored. Instead, add another column
with an indicator variable (for example, `"trusted"`, with values
`TRUE` or `FALSE`).

Here's an example in which a cell with a suspicious entry is highlighted.

```{r highlight_outlier}
mat <- data.frame(id=101:107,
                  date=c("2015-06-14", "", "2015-06-18", "", "", "2015-06-20", ""),
                  glucose=myround(c(149.3, 95.3, 97.5, 1.08, 108.0, 149.0, 169.4), 1),
                  stringsAsFactors=FALSE)
mat$date[2] <- mat$date[1]
mat$date[4:5] <- mat$date[3]
mat$date[7] <- mat$date[6]

hilitcells <- "C5"

excel_fig(mat, hilitcells=hilitcells)
```

It would be better to include an additional column that indicates
the outliers. The highlighting is nice visually, but it's hard to grab
that information for use in the later analysis.

```{r outlier_column, fig.width=5/4*5}
mat$outlier <- c(FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE)
excel_fig(mat)
```

Here's an example with males highlighted in blue and females in
pink. Rather than use highlighting to indicate sex, it's better to
include a `sex` column, with values `Male` or `Female`.

```{r highlight_sex}
mat <- data.frame(id=101:107,
                  date=c("2015-06-14", "", "2015-06-18", "", "", "2015-06-20", ""),
                  glucose=myround(c(149.3, 95.3, 97.5, 117.0, 108.0, 149.0, 169.4), 1),
                  stringsAsFactors=FALSE)
mat$date[2] <- mat$date[1]
mat$date[4:5] <- mat$date[3]
mat$date[7] <- mat$date[6]

hilitcells <- c(paste0(rep(LETTERS[1:3], each=3), 1+c(2,3,6)),
                paste0(rep(LETTERS[1:3], each=4), 1+c(1,4,5,7)))
hilitcolor <- c(rep("#F0DCDB", 3*3), rep("#DDE6F1", 4*3))

excel_fig(mat, hilitcells=hilitcells, hilitcolor=hilitcolor)
```

---

Next up: [Choose good names for things](names.html).

(Previous: [No calculations in the raw data files](no_calculations.html).)
